Future challenges for Research Software Engineering

### 2025-04-16 future_challenges_for_RSEng.tex

\subsection{Ethics and social consideration}
FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?

@@

- Corporate dominance:
  - AI and coding platforms might be increasingly controlled by a small number of large corporations.
  - RSE community might need to develop or advocate for open-source alternatives.

\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
\subsubsection{What do we want to have achieved}
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see, that researchers naturally accept them as part of their used services.

\subsubsection{Which challenges will still persist}
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with succesful RSE Teams.
On a more organisational level, we expect the sharing of models, to fit the different organizational constraints.

What are open questions here?
Organizations will have to come up with models for hiring people. While
a standardized RSE masters makes it easier, it will be interesting to see, if
RSE are still often hired inhouse or if it changes.

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
FG-TEXT-START: The RSE is probably one of the first professionalizations in the academic midlevel that is able
to collectively voice their concerns. This enables the possibility to reshape universities.
If more people are employed at the midlevel, this makes a steadier pool of people
available that can integrate new people that are regularly coming into departments and know the local facilities.
On the other hand an upcoming challenge can be, how universities decide to design the contracts of people employed
in RSE units. will they be permanent, or will there also be an up-or-out mentality?
FG-TEXT-END

FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.

>>>>

\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
- Emerging structures:
  - Universities/Institutes with central pools of RSEs, each specializing in data management, visualization, HPC, etc.
  - Potential for closer collaboration between domain scientists and RSEs to share workload effectively.

- Challenges:
  - Budget and administrative hurdles to creating permanent, well-funded RSE teams.
  - Aligning expectations: researchers might not realize that specialized RSE services exist (and are beneficial).

- Networking:
  - Need to share best practices among institutions with successful RSE teams.
  - Encourage knowledge exchange to find models that suit different organizational constraints.

\subsection{Ethics and social consideration}
The ethics and values of the Research Software Engineer are grounded in their practice.
They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
Sometimes these values are in direct conflict with the aims of their institution.
For example, their PI might insist on quickly developing some software without adhering to current best practices.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.

RSEs are publicly funded.
Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.

A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines.

\begin{FramedParagraphWithFootnotes}  
FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?

@@

- Corporate dominance:
  - AI and coding platforms might be increasingly controlled by a small number of large corporations.
  - RSE community might need to develop or advocate for open-source alternatives.
\end{FramedParagraphWithFootnotes}

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
